Compilation and Backend-Independent Vectorization for Multi-Party Computation

### 2022-04-28 overleaf/introduction.tex
Edits to 1 and 2.

We focus on the MOTION framework as our back-end for several reasons. First, it is the state of the art in terms of performance~\cite{Braun:2022}. Second, it provides an API over efficient implementation for a wide variety of cryptographic operations in three different protocols ---  Arithmetic GMW, Boolean GMW, and BMR --- which allows for protocol mixing~\cite{Ishaq:2019, Fang:2022}, a known backend-independent optimization. Third, MOTION provides API for SIMD-level operations, which amortize cost and lead to significant improvement in memory footprint and
throughput~\cite{NDSS:DemSchZoh15, CCS:ABFKLO18, Braun:2022}. It also enables MPC Source-level optimizations such SIMD vectorization.

Our second contribution is an analytical model for cost estimation of amortized schedules. Originally, we hoped that optimal scheduling (under our model) was tractable, as the problem
appeared simpler than the classical scheduling problem. Unfortunately, we show that optimal scheduling is NP-hard via a reduction to the Shortest Common Supersequence (SCS) problem.
Cost estimation is important as it drives not only vectorization but other optimizations such as protocol mixing optimizations and scheduling as well~\cite{Ishaq:2019, Fang:2022}.

@@

Our results emphasize the importance of backend-independent optimizations --- vectorization (described in this work) and protocol mixing (tackled in previous works~\cite{CCS:BDKKS18,Ishaq:2019, Fang:2022}) are two optimizations readily available at the level of MPC Source. We believe that our work can lead to future work on backend-independent compilation and optimization, bringing in new MPC optimizations and combinations of optimizations, much in the vein of standard compilers, and bring MPC programming technology closer to practice and wider applicability.

@@

The rest of the paper is organized as follows. \secref{sec:overview} presents an overview of the compiler. \secref{sec:model} describes our model for cost estimation and argues NP-hardness of optimal scheduling and \secref{sec:compiler} describes the compiler. \secref{sec:experiments} presents implementation and evaluation. \secref{sec:related} discusses related work and \secref{sec:conclusion} concludes.

>>>>

We focus on the MOTION framework as our back-end for several reasons. First, it is the state of the art in terms of performance~\cite{Braun:2022}. Second, it provides an API over efficient implementation for a wide variety of cryptographic operations in three different protocols ---  Arithmetic GMW, Boolean GMW, and BMR --- which allows for protocol mixing~\cite{Ishaq:2019, Fang:2022}, a known backend-independent optimization. Third, MOTION provides API for SIMD-level operations, which amortize cost and lead to significant improvement in memory footprint and
throughput~\cite{NDSS:DemSchZoh15, CCS:ABFKLO18, Braun:2022}. It enables MPC Source-level vectorization, a key focus of this paper.

Our second contribution is an analytical model for cost estimation of amortized schedules. Originally, we hoped that optimal scheduling (under our model) was tractable, as the problem
appeared simpler than the classical scheduling problem. Unfortunately, we show that optimal scheduling is NP-hard via a reduction to the Shortest Common Supersequence (SCS) problem.
Cost modeling is important as it drives not only vectorization but optimizations such as protocol mixing and scheduling as well~\cite{Ishaq:2019, Fang:2022}.

@@

Our results emphasize the importance of backend-independent optimizations --- vectorization (described in this work) and protocol mixing (tackled in previous works~\cite{CCS:BDKKS18,Ishaq:2019, Fang:2022}) are two optimizations readily available at the level of MPC Source. We believe that our work can lead to future work on backend-independent compilation and optimization, ushering new MPC optimizations and combinations of optimizations in the vein of standard compilers, and thus bringing MPC programming technology closer to practice and wider applicability.

@@

The rest of the paper is organized as follows. Section \secref{sec:overview} presents an overview of the compiler. Section \secref{sec:model} describes our model for cost estimation and argues NP-hardness of optimal scheduling. Section \secref{sec:compiler} details the phases of the compiler and Section \secref{sec:vectorization} focuses in on vectorization. \secref{sec:results} presents the experimental evaluation. Section \secref{sec:related} discusses related work and Section \secref{sec:conclusion} concludes.

### 2022-04-28 overleaf/overview.tex
Edits to 1 and 2.

As a running example, consider Biometric matching, a standard MPC benchmark.
An intuitive (and naive) implementation is as shown in Listing~\ref{tab:source_and_MPC_source_and_optimized_MPC_source}(a).
Array \python{C} is the feature vector of \texttt{D} features that we wish to match and \python{S}
is the database of \python{N} size-\python{D} vectors that we match against.

Our compiler takes essentially standard IMP~\ana{Need citation here.}
syntax and imposes certain semantic restrictions. (We detail the restrictions in the following sections.)
The programmer writes an iterative program and annotate certain inputs
and outputs as \emph{shared}. In the example arrays \python{C} and \python{S}
are shared, meaning that they store shares, however, the array sizes \python{D} and
\python{N} respectively are plaintext. The code iterates over the entries in the
database and computes the Euclidean distance (its square actually) of the current
entry \python{S[i]} and \python{C}. The program returns the index of the vector that gives
the best match plus the corresponding sum of squares.

@@

Our compiler generates an intermediate representation, MPC Source. MPC Source
for Biometric Matching is shown in
Listing~\ref{tab:source_and_MPC_source_and_optimized_MPC_source}(b). First, MPC Source is an SSA form.
Second, it is linear. The conditional in lines 13-14 in IMP Source turns into the linear code in lines 12-16 in MPC Source.
The test turns into a CMP operation (\python{t = CMP(sum!3,min\_sum!2)}), followed by the
true-branch sequence, followed by the MUX operations. The first MUX operation selects the value
of \python{min\_sum}: if \python{t} is true, then \python{min\_sum} gets the value of the second multiplexer
argument,  \python{min\_sum!3}, otherwise it takes the value of the third argument, \python{min\_sum!2}.
Third, MPC Source is a special form of SSA. The SSA $\phi$-nodes at the if-then-else (lines 13-15) turn into
MUX operations, while the $\phi$-nodes at for loops turn into what we call pseudo PHI nodes with a straightforward semantics.

We turn to our analytical model to compute the \emph{cost} of the program. Assuming fixed
cost $\beta$ for a local MPC operation (essentially just ADD) and cost $\alpha$ for a remote
MPC operation (e.g., MUX, CMP, and remaining operations), the cost of the iterative schedule
will be $ND(2\alpha+\beta) + N(3\alpha$).

A key contribution is the vectorizing transformation. We can compute all $N*D$
subtraction operations (line 9) in a single SIMD instruction; similarly we can compute
all multiplication operations (line 10) in a single SIMD instruction. And while we cannot
vectorize computation of the N individual sums, we can compute the N sums in parallel.
Our compiler \emph{automatically detects these opportunities and transforms the program}.
It is standard that MPC researchers write vectorized versions of the Biometric
program by hand; we are the first (to the best of our knowledge) to automatically
transform a naive, iterative MPC program into an unintuitive vectorized one.

@@

The compiler determines that SUB and MUL in ``naive'' MPC Source (lines 9 and 10 in (b))
can be fully vectorized into the SIMD SUB and MUL in optimized MPC Source (lines 9 and 10 in (c)).
Notation \python{p[I,J]} denotes a 2-dimensional array with fully vectorized dimensions.
The computation of sum (line 11 in (b))
is sequential across the $j$-dimension, but it is parallel across the $i$-dimension.
The loop in lines 12-16 in (c) illustrates; here \python{p[I,j]} refers to the $j$-th column in \python{p}.
Unfortunately, CMP and MUX remain sequential.

In MPC backends, executing $n$ operations ``at once'' in a single SIMD operation costs a lot less than executing those $n$ operations one by one.
This is particularly important when there is communication (i.e., in remote), since many 1-bit values are sent at once rather than sequentially.
%In MPC compilers a vectorized operation that computing M operations "at once" costs
%essentially the same ($\alpha$ or $\beta$) as an individual operation.
We elaborate on the cost model in the following section but for now consider that
each operation has a \emph{fixed} portion (does benefits from amortization) and
a \emph{variable} portion (does not benefit from amortization): $\alpha = \alpha_\mathit{fix} + \alpha_\mathit{var}$.
This gives rise to the following formula for amortized cost: $f(n) = \alpha_\mathit{fix} + n\alpha_\mathit{var}$,
as opposed to unamortized cost $g(n) = n\alpha_\mathit{fix} + n\alpha_\mathit{var}$. (We extend the same reasoning to
$\beta$-instructions.)

Thus, the fixed cost of the vectorized program amounts to
$2\alpha_\mathit{fix}$ + $D \beta_\mathit{fix} + N(3\alpha_\mathit{fix})$. (The variable cost is the same in both the vectorized and non-vectorized programs.)
The first term in the sum corresponds to the vectorized
subtraction and multiplication, the second term corresponds to the FOR loop on $j$ and the third
one corresponds to the remaining FOR loops on $i$. Clearly, $2\alpha_\mathit{fix} + D\beta_\mathit{fix} + N(3\alpha_\mathit{fix}) << ND(2\alpha_\mathit{fix}+\beta_\mathit{fix}) + N3\alpha_\mathit{fix}$.
Empirically, we observe 10x to 50x improvement over un-vectorized Biometric Matching in circuit generation time and setup time and 5x to 30x improvement in online time.
Additionally, the un-vectorized version runs out of memory for $N=256$, while the vectorized one runs with the standard maximal input size $N=4,096$.
\ana{Edit numbers with final experiments.}

>>>>

As a running example, consider Biometric matching, a standard MPC benchmark.
An intuitive (and naive) implementation is as shown in Listing~\ref{tab:source_and_MPC_source_and_optimized_MPC_source}(a).
Array {\sf C} is the feature vector of {\sf D} features that we wish to match and {\sf S}
is the database of {\sf N} size-{\sf D} vectors that we match against.

Our compiler takes essentially standard IMP~\ana{Need citation here.}
syntax and imposes certain semantic restrictions. (We detail the restrictions in the following sections.)
The programmer writes an iterative program and annotates certain inputs
and outputs as \emph{shared}. In the example arrays {\sf C} and {\sf S}
are \texttt{shared}, meaning that they store shares, however, the array sizes {\sf D} and
{\sf N} respectively are plaintext. The code iterates over the entries in the
database and computes the Euclidean distance of the current
entry {\sf S[i]} and {\sf C} (its square actually). The program returns the index of the vector that gives
the best match plus the corresponding sum of squares.

@@

Our compiler generates an intermediate representation, MPC Source. MPC Source
for Biometric Matching is shown in
Listing~\ref{tab:source_and_MPC_source_and_optimized_MPC_source}(b). First, MPC Source is an SSA form.
Second, it is linear. The conditional in lines 13-14 in IMP Source turns into the linear code in lines 12-16 in MPC Source.
The test turns into the CMP operation {\sf t = CMP(sum!3,min\_sum!2)}, followed by the
true-branch sequence, followed by the MUX operations. The first MUX operation selects the value
of {\sf min\_sum}: if {\sf t} is true, then {\sf min\_sum} gets the value of the second multiplexer
argument,  {\sf min\_sum!3}, otherwise it takes the value of the third argument, {\sf min\_sum!2}.
Third, MPC Source is a special form of SSA. The SSA $\phi$-nodes at the if-then-else (lines 13-15) turn into
MUX operations, while the $\phi$-nodes at for-loops turn into \emph{pseudo} PHI nodes with a straightforward semantics.

We turn to our analytical model to compute the \emph{cost} of the iterative program. Assuming
cost $\beta$ for a local MPC operation (essentially just ADD) and cost $\alpha$ for a remote
MPC operation (e.g., MUX, CMP, and remaining operations), the cost of the iterative schedule
will be $ND(2\alpha+\beta) + N(3\alpha$).

A key contribution is the vectorizing transformation. We can compute all $N*D$
subtraction operations (line 9 in (b)) in a single SIMD instruction; similarly we can compute
all multiplication operations (line 10) in a single SIMD instruction. And while computation
of an individual sum remains sequential, we can compute the $N$ sums in parallel.
Our compiler \emph{automatically detects these opportunities and transforms the program}.
It is standard that MPC researchers write vectorized versions of the Biometric
program by hand; we are the first (to the best of our knowledge) to automatically
transform a naive, iterative MPC program into an unintuitive vectorized one.

@@

The compiler determines that SUB and MUL in ``naive'' MPC Source (lines 9 and 10 in (b))
can be fully vectorized into the SIMD SUB and MUL in optimized MPC Source (lines 9 and 10 in (c)).
Notation {\sf p[I,J]} denotes a 2-dimensional array with fully vectorized dimensions.
The computation of sum (line 11 in (b))
is sequential across the $j$-dimension, but it is parallel across the $i$-dimension.
The loop in lines 12-16 in (c) illustrates; here {\sf p[I,j]} refers to the $j$-th column in {\sf p}.
Unfortunately, CMP and MUX remain sequential.

In MPC backends, executing $n$ operations ``at once'' in a single SIMD operation costs a lot less than executing those $n$ operations one by one.
This is particularly important when there is communication (i.e., in remote), since many 1-bit values are sent at once rather than sequentially.
%In MPC compilers a vectorized operation that computing M operations "at once" costs
%essentially the same ($\alpha$ or $\beta$) as an individual operation.
We elaborate on the cost model in Section~\secref{sec:model} but for now consider that
each operation has a \emph{fixed} portion (does benefits from amortization) and
a \emph{variable} portion (does not benefit from amortization): $\alpha = \alpha_\mathit{fix} + \alpha_\mathit{var}$.
This gives rise to the following formula for amortized cost: $f(n) = \alpha_\mathit{fix} + n\alpha_\mathit{var}$,
as opposed to unamortized cost $g(n) = n\alpha_\mathit{fix} + n\alpha_\mathit{var}$. (We extend the same reasoning to
$\beta$-instructions.)

Thus, the fixed cost of the vectorized program amounts to
$2\alpha_\mathit{fix}$ + $D \beta_\mathit{fix} + N(3\alpha_\mathit{fix})$. (The variable cost is the same in both the vectorized and non-vectorized programs.)
The first term in the sum corresponds to the vectorized
subtraction and multiplication (lines 9-10 in (c)), the second term corresponds to the FOR loop on $j$ (lines 12-16) and the third
one corresponds to the remaining FOR loops on $i$ (lines 19-25). 
Clearly, $2\alpha_\mathit{fix} + D\beta_\mathit{fix} + N(3\alpha_\mathit{fix}) << ND(2\alpha_\mathit{fix}+\beta_\mathit{fix}) + N3\alpha_\mathit{fix}$.
Empirically, we observe 10x to 50x improvement over un-vectorized Biometric Matching in circuit generation time and setup time and 5x to 30x improvement in online time.
Additionally, the un-vectorized version runs out of memory for $N=256$, while the vectorized one runs with the standard maximal input size $N=4,096$.
\ana{Edit numbers with final experiments.}
